---
title: "Functional Programming"
output: learnr::tutorial
runtime: shiny_prerendered
description: "Learn about functions and loops"
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


# Functional programming

Functions and loops

## Functions

Functions form the backbone of R. Functions are pieces of code that take an input, processes it and produces an output. R contains a set of built in functions for doing basic things like generating a sequence of numbers between 1 and 10. Try it yourself:

```{r seq, exercise = TRUE}
seq(1, 10)
```
Functions in R have 4 general components: a name, inputs, what happens to the inputs and the output taking the general form `function_name(input)`

In the case above, the function is called `seq`, the inputs are `1` and `10` indicating the first and last numbers of the sequence respectively and the output is a sequence of 10 numbers.  

Functions can be nested within each other - this is the functional programming aspect of R. Then the output of one function is sent directly to the next function.  

For example we can take the mean of our sequence of numbers above:  
```{r meanseq, exercise = TRUE}
mean(seq(1,10))
```

When functions are nested like this, R does not save the intermediate output (here, the sequence of numbers) to the environment.

*But we know this already* you say. Yeah, I say, but manual functions behave the same way so you need to understand the basics of R before you move on to more advanced stuff.

***

### Manual functions

Built in functions are handy but when we want to do more complex tasks, we need to make our own functions. This is also useful when we need to do repetitive tasks. We could copy the code over and over again while changing parameter values but this increases the chance of making an error and is harder to debug or modify.

> A guide to functional programming in R is Hadley Wickham's book [Advanced R](http://adv-r.had.co.nz/Functional-programming.html). Refer to it and other for more expertise

Manual functions in R take a general structure following what we've described above:

```
function_name <- function(x) {Function does stuff}
```

* `function_name` is the name of the function. This is what is used to call the function when it is stored in the R environment
* `function()` tells R that this is a manual function
* `x` are the user inputs of the function
* `{}` tell R what to do with `x`. This can be as many things as you want. 

Once the function is stored in R's memory, we can use it like any regular function in R

***

### Your first function

Let's make a function to print us a message telling us what the mean of a vector of numbers is. Complete the code below to print the mean of our sequence of numbers between 1 and 10:
```{r first, exercise = TRUE}
print_mean <- function(x){
  message(paste("The mean is", mean(x)))
}
print_mean()
```

The answer should be 5.5  

That's a small example of printing strings. But strings aren't useful for further processing. what if we *also* wanted to use the calculated mean? We'd have to tell R to calculate the mean separately and provide the output in a way that can be used in further code. We can ask the function to calculate an intermediary variable called `x_mean` *and* return that value to the R environment.

Run this modified function and see what the difference is:
```{r second, exercise = TRUE, exercise.lines = 8}
print_mean <- function(x){
  x_mean <- mean(x)
  message(paste("The mean is", x_mean))
  return(x_mean)
}

```
```{r second-hint}
don't forget to run the function using print_mean() with the sequence of numbers we want
```

We've made three modifications to our original `print_mean` function. Let's break this down:

1. When the function gets a sequence of numbers (called x) it's going to calculate the mean of those numbers and store them in a variable called `x_mean`. `x_mean` doesn't get stored in the environment and only exists while the function is running. 
2. After `x_mean` is calculated the function is going to print a message saying what the value stored in `x_mean` is. So the calculation of the mean is separated from the printing of the message
3. To tell R to inform us of what `x_mean` is we tell it to `return` 

Now the function tells us what the mean is and we can either assign that to a variable name or feed the output into another function

> Functions can take more than one input. They can also be assigned "default" values to parameters that get overwritten by any user input.
> However, functions can only return one output using `return`. To return multiple values, we need to tell R to save them to a list or a data frame.

You don't have to use `return` - it will also work if you jus printed `x_mean` but it is best practice as it specifically tells R what value to give. If you don't tell R to produce an output then it won't show any results. Try it - modify the code above by removing the `return` function and see what happens. 

### Multiple returns

Say we want a function to give us more information at the end. We can return multiple values as a list:
```{r multi, exercise = TRUE, exercise.lines = 15}
print_mean <- function(x){
  x_mean <- mean(x)
  message(paste("The mean is", x_mean))
  mean_TF <- ifelse(x_mean < 4, TRUE, FALSE)
  results <- list(x_mean = x_mean, mean_TF = mean_TF)
  return(results)
}

fun_results <- print_mean(seq(1,10))
fun_results
```

What happened when you ran the above?

There are two more modification to this third function compared to the second version.

1. There's a named variable called `mean_TF` which is an `ifelse` statement - that is, a function nested within a function. This `ifelse` statement is *vectorised* which is something we prefer over `if` and `else` statements because of the way that R processes objects. Here the statement is testing whether the logic that `x_mean` is less than 4 and returning a logical variable (`TRUE/FALSE`)
2. `results` is a named variable containing a list with two items: our `x_mean` variable and our `mean_TF` variable. This is what is returned at the end

To help us see the results we've saved the output (the list) into a variable called `fun_results`

### Recap

**That's the basics**

```{r fun-recap}
quiz(caption = "Test your knowledge",
  question("What is the structure of a manually defined function taking the input x?",
           answer("function(x){}", correct = TRUE),
           answer("function(x)"),
           answer("function{x}"),
           answer("function{x}()"),
           allow_retry = TRUE,
           random_answer_order = TRUE
  ),
  question("How do you return multiple values from a function?",
           answer("Save your output as a named list or data frame", correct = TRUE),
           answer("Use return on a single variable"),
           answer("Add more variable to the brackets in function()"),
           allow_retry = TRUE,
           random_answer_order = TRUE
  ),
  question_text("Enter the output of the function below if a was 22",
                answer("32", correct = TRUE),
                allow_retry = TRUE
  )
)
```

```
function(a){
  b <- a + 10
  c <- b - 12
  return(b)
}
```


## Loops

Sometimes we need to run a function multiple times while changing the inputs. This is where loops come in. There are a few different types of loops but we will only look at one: the `for` loop.

> `for` loops take the general structure `for(a condition){do stuff}`

The logic is that R will repeatedly do whatever the loop says as long as the condition is met. It's easier to show:

```{r loop1, exercise = TRUE}
month <- month.name
head(month)
for(i in seq_along(month)){
  print(i)
}
```
What do you think this loop is doing?

* `month` is a character vector containing the names of the months
* `seq_along` is a useful function for telling which observation you are at for a vector - try it out on its own. In other words, moving along the vector and identifying the n^th^ observation
* The condition of the for loop is `i in seq_along(month)`. `i` is a placeholder indicating in n^th^ observation (or i^th^ observation to be exact). This can be called anything you want but `i` is from mathematical notation.
* `print` is showing us the value of `i`

So in words the loop is saying *at each observation from the start of month vector to the end of the month vector, tell me how far along the month vector you are*. In other words, which number of the month is it?

Printing is pretty basic. Let's use the loop to manipulate our vector

```{r loop2, exercise = TRUE}
month <- month.name
month.abb
for(i in seq_along(month)){
  month[i] <- month.abb[i]
  print(month)
}
```

There are two changes to this loop:

1. ` month[i] <- month.abb[i]` where `month[i] <-` says assign what's on the right of the arrow to the i^th^ observation in the vector `month` & `month.abb[i]` is the i^th^ observation of the `month.abb` vector, which is a built in vector of the months in an abbreviated format.
    * Thus, replace the i^th^ observation in `month` with the equivalent observation in `month.abb`
2. `print(month)` says print our updated `month` vector every time we update it. The changes should appear sequentially

Check your Environment tab, there should be a vector called `month` that has the abbreviated months.  

**What would happen if we didn't have `[i]` next to `month`?** Try it.

> If you don't explicitly assign new values to a vector then R will only save the last object produced

### Your turn

The loop below manipulates the `test` vector by dividing the i^th^ observation by 5, then reporting the cumulative sum at each iteration & saving the cumulative sum to `test`. But there are two errors in the loop. Find the errors and fix it.
```{r test, exercise = TRUE, exercise.lines = 6}
test <- seq(10,50, 5)

for(i in seq_along(test)){
  test[i] <- test[i]/5
  print(test[i])
  test <- sum(test[i:1])
}
```
```{r test-hint}
work through what each step is doing and how it matches/differs from what you want it to do
```
```{r test-solution}
test <- seq(10,50, 5)
for(i in seq_along(test)){
  test[i] <- test[i]/5
  print(test)
  test[i] <- sum(test[1:i])
}
```

***

### Alternative structures

`seq_along` is useful in most cases and it saves writing `for(i in 1:length(vector))` but sometimes we want to work with the actual value stored in the i^th^ observation. Then, we need to tweak our for loop.  

This loop uses *two* placement holders: `i` and `n`. It's not the most elegant of loops or the best way to do so but it does show how loops work. See if you can figure out what is happening here
```{r for, echo = TRUE, exercise = TRUE}
month <- seq(12,1)
i <- 1
for(n in month){
  month[i] <- month.name[n]
  i <- i + 1
}
month
```
The function is converting month number into month names and the month numbers are in reverse order. Thus using `seq_along` wouldn't work because that works in increasing order. Now we are using the actual values stored in `month` rather than the observation number, this wouldn't work for a character vector.

`i` works as before and tracks the observation number in the vector. `n` is an additional place holder that stores the actual value in `month`.

Another way of writing this loop is to save the actual value in the observation into a variable:
```{r for2, echo = TRUE, exercise = TRUE}
month <- seq(12,1)
month_name <- array()
for(i in seq_along(month)){
  month_name[i] <- month.name[month[i]]
}
month_name
```

`month_name` is an empty variable that we fill using the loop.

***

## If and ifelse statements

If and ifelse statements are fundamental programming structures that replicate logical if statements. These statements follow the general structure:

> if(condition){do something}

You'll notice that it's similar to a function. If the condition isn't met then the if statement doesn't do anything

Ifelse is an extension that has an followup logical statement based on a true and false.

> ifelse(condition, do something if condition is true, do something if condition is false)

Let's try it.  

Run the following code and see what happens. Then try for different values of `num`. What is the output if `num` is greater than 20?
```{r if, exercise = TRUE}
num <- 10
if(num <20){
  print("The number is less than 20")
}
```

Now let's try an ifelse statement. What happens with different numbers?
```{r ifelse, exercise = TRUE}
num <- 10
ifelse(num < 20, # The condition
       paste("The number is less than 20"), # do this if the condition is true
       paste("The number is greater than 20")) # do this if the condition is false
```

***

## An example: Predator-Prey interactions

When are loops and functions most useful? When they can be integrated in a program. We will use a simulation of predator-prey interactions as an example. 

Predator-prey interactions simulate how many prey a predator can capture after a specific amount of time and for a given density of prey. We use this example with undergraduate biology students to demonstrate statistical modelling, experimental design and collecting data.

Normally we would get students to do this laboratory practical in class by picking up counters and putting them in jars while blindfolded. We can also see whether the use of different types of jars affects the efficiency of the predator. This is the experimental design:

* Response variable - Number of prey caught (`Ha`)
* Total foraging time - 1 minute, a constant (`T`)
* Predictor variables:
    * Prey density - user defined treatments (`H`)
    * Type of jar used - jar with a lid or no lid (`yes` or `no`)

Normally the students will collect data to parameterise the functional response. Instead I've created a function that will predict new values of prey captured using a functional response formula that is already parameterised.

Let's look at the function:

***

### An interactive function

R has some capacity to be interactive. It can ask a user to input variables.
The funtion `functional_response` will return the number of prey captured for a given prey density and type of jar used. The function will ask for these two inputs each time the function is run. Look at the code below and try to understand the different components. Then try running the code yourself with different inputs.

```{r interaction, echo = TRUE}
#### Interactive function
# Simulation of the predator-prey functional response - DO NOT CHANGE
# To run: click Source (cmd or ctrl+shift+S) or Run All (ctrl+alt+r)
# Or source("<insert file location here>", echo = FALSE)

functional_response <- function(){
  
  # Introduce the simulation to the user - prints a message
  cat(paste(
    "",
    "You have some counters (prey), a piece of A4 paper & a jar.",
    "You spread the counters randomly on the A4 sheet.",
    sep="\n"))
  
  # Ask the user for the prey density for the functional response
  prey_density  <- readline("How many prey counters are used? ") # ask for prey density
  
  # Check the user has inputted a number properly
  if(!grepl("^[0-9]+$", prey_density)){ # check whether the input contains numeric characters between 0-9 using regex (regular expressions)
    message("Please enter an integer") # If the input is not a number, tell them to do it again
    return(functional_response()) # Return to the beginning of the function and start again
  }
  prey_density <- as.integer(prey_density) # If the prey density input is a number, turn it into an interger
  
  # Ask the user for the type of jar used
  cat(paste(
    "",
    "Every second you (the predator) tap the sheet to find and pick up a prey counter while blindfolded.",
    "You have 1 minute to put as many prey counters as you can in the jar.",
    "There are two types of jars you can use while handling prey.",
    "Enter 1 to use a jar with a lid that you have to open and close.",
    "Enter 2 to use a jar without a lid.",
    "",
    "What type of jar is used?",
    sep="\n"))
  lid_used <- menu(c("Lid", "No Lid"))
  
  # Calculate the number of prey caught (the functional response) based on the user defined input above
  if(lid_used != 0){ # Check that the use has chosen the jar used properly (1 or 2)
    
    # Use this model if using a jar with a lid
    if(lid_used == 1){
      Ha <- (0.2 * prey_density)/(1 + 0.2 * 0.03 * prey_density) 
    }
    
    # Use this model if using a jar without a lid
    if(lid_used == 2){
      Ha <- (0.7 * prey_density)/(1 + 0.7 * 0.05 * prey_density) 
    }
    
    # Add in some variation around the predicted value so that users don't get the exact parameterised functional response
    Ha <- Ha + sample(seq(-3,3), 1) 
    
    # Make sure there are no negative prey items caught!
    if(Ha < 0){
      Ha <- 0 # Make prey caught 0 if less than 0
    }
    
    # Make sure the number of prey caught doesn't exceed the number of prey available!
    if(Ha > prey_density){
      Ha <- prey_density # If prey caught is greater than the number of prey available, make it the maximum possible
    }
    
    # Print a message showing the results
    message("The number of prey caught is ", floor(Ha), ". Well done!") 
  }
}

# Actually run the function and tell R that it's interactive 
if(interactive()) functional_response()
```

Note:

* The function doesn't have any inputs in `function()` because it will ask the user for them each time
* `readline` is the function to ask for a single user input
* `menu` is the function to ask the user to chose from a number of options
    * Here the option is press 1 to use a jar with a lid or press 2 to use a jar without a lid
* There are two parameterised functional responses - one for a jar with a lid and one for a jar without a lid
* `floor` is a function to round the number of prey caught to the lowest whole number

In the practical, students will need to run the above function for 10 prey densities, repeated 3 times, for both jar treatments - so 60 times in total. But we don't have to do that manually - that is what loops are for!

***

### Functions and loops

Here is a non-interactive version of the function above. It doesn't have the printed messages asking for user input. This time, the function needs 3 inputs as indicated by `function(prey_density, lid_used, total_time)`: the prey density used, the type of jar used and the total foraging time, respectively.

Run the code chunk to load the function into the R environment:

```{r functional-response, echo=TRUE}
# Functional response function
functional_response <- function(prey_density, lid_used, total_time){
  
  # Check jar type is inputted correctly
  if (!lid_used %in% c("yes", "no")) {
    stop("Lid used is not inputted correctly. Use 'yes' or 'no' in all lowercase")
  }
  if(lid_used == "yes"){
    Ha <- (0.2 * prey_density * total_time)/(1 + 0.2 * 0.03 * prey_density)
  }
  if(lid_used == "no"){
    Ha <- (0.7 * prey_density * total_time)/(1 + 0.7 * 0.05 * prey_density)
  }
  Ha <- Ha + sample(seq(-3,3), 1)
  if(Ha < 0){
    Ha <- 0
  }
  if(Ha > prey_density){
    Ha <- prey_density
  }
  # message("The number of prey caught is ", floor(Ha))
return(floor(Ha))
  }
```

The function will check that the character vector indicating the type of jar to be used is correct because R is case sensitive. The function will return the number of prey caught `Ha` as indicated by `return(Ha)`. The function `return` specifically tells R to tell us the output, otherwise R will keep it to itself! Only one output is allowed (unless extra steps are taken).
The rest of the function is the same.

***

### Using the functional response function in a loop

Now let's use a loop to do our entire experiment in one go! No manual inputs for us. In fact we can do as many treatments or replicates as we want. Let's do prey densities between 5 and 100 in increments of 5 and 3 replicates. Since the function inputs are required we can set them up in the environment for the function:

```{r parameters, echo = TRUE}
# Set parameters for the function
total_time <- 1 # total foraging time in minutes
no_treatments <- seq(from = 5, to = 100, by = 5) # prey density treatments between 5 and 100
replications <- 3 # number of replications

# a numeric vector of prey densities for all treatments, jar types and replications 
prey_density <- rep(rep(no_treatments, replications),2) # repeated twice for each jar type

# a character vector of the jar type
lid_used <- sort(rep(c("no", "yes"), length(prey_density)/2)) # "yes" or "no"
```

By setting up the parameters outside the function or loop, we can easily modify the parameters of the function and feed the new values into the loop. This helps us debug and is cleaner and easier to read.

Now for the actual loop itself. We need to be able to store the output of the looped function.
Lists are the fastest way to do so in R because R is designed for lists and matrices.

```{r loop, echo = TRUE}
# Create an empty list called prey_caught to populate 
prey_caught <- list()

for(i in seq_along(prey_density)){
  # Run the functional response for the nth observation in the vector prey_density representing all our observations and save it to the list
  prey_caught[[i]] <- functional_response(prey_density = prey_density[i],
                                         lid_used = lid_used[i],
                                         total_time = total_time)
  
  # Prepare the list for further analysis
  prey_caught[[i]] <-  cbind(prey_caught[[i]], prey_density[i]) # add a column for the prey density used to calculate the number of prey caught for that observation (row)
  prey_caught[[i]] <-  cbind(prey_caught[[i]], 1/prey_caught[[i]]) # add a column for the inverse of the number of prey caught - for statiscally parameterising the functional response
  prey_caught[[i]][,3] <-  ifelse(prey_caught[[i]][,3] == Inf, 0, prey_caught[[i]][,3]) # Housekeeping - turn undefined values of prey caught (from calculating 1 divided by 0) into 0. If the number of prey caught was 0
}

# Turn our list into a data frame
prey_caught <- data.frame(do.call("rbind", prey_caught))

# Label the columns
colnames(prey_caught) <- c("Ha", "H","Ha.1", "HT.1") # ".1 is inversed columns"

# Add type of lid used to our data frame
prey_caught$lid_used <- lid_used
```

Remember:

* Here we are using a `for` loop with the n^th^ observation denoted `i`. This can be called anything you want but `i` is from mathematical notation.
* `seq_along` is a useful function for telling which observation you are at for a vector.

Things to note:

* Lists can only contain one variable type, unlike a data frame, but that's why they are fast and memory efficient for high performance computing
    * If we were to add the jar type (a character vector) to the list within the loop, then all our numeric output will be converted to characters (default R behaviour) - not what we want!
* `do.call` is a handy function for lists. It collapses multidimensional lists into one dimension, here appending them by rows (i.e. adding new observations by rows at the end)

***

### Visualising the data

Now we can plot our results and conduct our linear regression:

```{r plot, message=FALSE, warning=FALSE, comment=NA}
library(tidyverse)

prey_caught %>% # Absolute values
  ggplot(aes(H, Ha, colour = lid_used)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, fullrange=TRUE) +
  labs(x = expression(paste("Prey density  (H)")),
       y = expression(paste("Prey captured  (Ha)")),
       colour = "Jar used") +
  theme_classic()

# Linear model
summary(lm(Ha.1 ~ HT.1 * lid_used, prey_caught))
```

That has saved us from running the code `r length(prey_density)` times! Now we can do it in one!

***

## Summary

A final tip about writing programs: if you can't explain it in words, then you can't code it. One concept in computer science is to explain your coding problems to a rubber duck. If you can explain your problem and desired outcome to a rubber duck then you can solve your problem.